Crunching Large Graphs with Commodity Processors

### 2011-04-20 doc/hotpar2011/softxmt-hotpar2011.tex
most of preston's changes

%\textheight 9in
%\textwidth 6.5in
%\columnsep 1pc
\renewcommand{\baselinestretch}{0.8}
%\renewcommand{\baselinestretch}{0}

@@

\authorinfo{Jacob Nelson$^{\dagger}$,
  Brandon Myers$^{\dagger}$,
  A.H. Hunter$^{\dagger}$,
  Preston Briggs$^{\dagger}$
  Luis Ceze$^{\dagger}$,
  Carl Ebeling$^{\dagger}$,\\
  Dan Grossman$^{\dagger}$,
  Simon Kahan$^{{\dagger \ddagger}}$,
  Mark Oskin$^{\dagger}$,
}{\textdagger University of Washington, \textdaggerdbl Pacific Northwest National Laboratory}{
  \{nelson, bdmyers, ahh, preston, luisceze, ebeling, djg, skahan, oskin\}@cs.washington.edu}

@@

\begin{abstract}
   Crunching large graphs is the basis of many emerging applications,
   such as social network analysis and bioinformatics. Graph analytics
   algorithms exhibit little locality and therefore present significant
   performance challenges. Hardware multithreading systems (e.g, Cray
   XMT) show that with enough concurrency, we can tolerate long
   latencies. Unfortunately, this solution is not available with
   commodity parts.

   Our goal is to develop a latency-tolerant system built out of
   commodity parts and mostly in software. The proposed system includes
   a runtime that supports a large number of lightweight contexts,
   full-bit synchronization and a memory manager that provides a
   high-latency but high-bandwidth global shared memory. This paper
   lays out the vision for our system, and justifies its feasibility
   with a performance analysis of the runtime for latency tolerance.

@@

@@

Many important emerging applications such as social network analysis,
bioinformatics, and sensor networks rely on crunching very large
graphs. Unfortunately, the computational cost of these applications
gets quickly out of hand. While tools to analyze social networks and
query semantic graphs with billions of entities exist
today~\cite{joslyn:hpsf, ediger:msda, brocheler:cosi, pegasus,
  pregel}, graphs of interest to defense applications are expected to
have trillions of entities~\cite{hartley:mssg, kolda:imkd}. Speeding
up these applications at a low cost would have a large impact in how
we deal with large data-sets and make them even more valuable.

@@

Multithreading is a technique that has been used successfully to
implement efficient computations for these
graphs~\cite{bader:bfsmta}. The Cray XMT is an example of such an
approach: it solves the memory latency problem through concurrency
rather than caching. Each XMT processor supports 128 hardware contexts
and 1024 outstanding memory operations, and is able to switch contexts
every cycle~\cite{tera, feo-xmt}. This ability comes at a cost: the
XMT is an expensive, non-commodity machine.

@@

We believe we can build a system based mostly on commodity parts that
can attain XMT-like performance with a familiar, XMT-like,
programming model, but at a fraction of the cost. Therefore, our goal
is to build a system that has good performance on low-locality graph
codes but is implemented with cheap commodity processors and simple
additional support in an FPGA. This approach has the added
benefit of not sacrificing general-purpose performance.

@@

Our goal is to preserve the XMT programming model: large shared global
address space, explicit concurrency with a large number of threads,
and full-bit synchronization.  We partition the overall address
space into a global shared address space and per node private address
spaces. Locality may be exploited directly by the programmer in the
node private address spaces, just as in a conventional cache-coherent
multiprocessor node. In contrast, locality cannot be exploited in the
global space: its value is in providing high bandwidth random access
to any word in a shared data structure by any processor.

@@

The model promises efficiency subject to the presence of sufficient
concurrency.  Exactly how that concurrency is expressed by the
programmer is language dependent. The goal of our system is to support
that concurrency and consequently provide latency tolerance on a large
address space. We do so via software multithreading: computation that
is about to execute a long latency operation (e.g. a remote memory
reference) initiates the operation and then yields to the scheduler,
which quickly resumes execution of another computation. Concurrency is
also used to make more efficient use of network bandwidth: it
enables aggregation of short memory requests into coarser requests,
which leads to large network messages and consequently better
bandwidth utilization.

@@

Synchronization on locations in the private address space of a node
works the the same way as on conventional systems.  Synchronization on
global addresses works differently: we provide atomic operations such
as {\tt int\_fetch\_add} as well as operations using full-bits, as on
the Cray XMT.  As is true for other long latency operations,
synchronization latency is tolerated by yielding to the scheduler.
Even non-blocking global atomics yield so the core can switch to
another computation while the synchronization is accomplished outside
of the pipeline.  In addition, the system can also exploit aggregation
when concurrency is abundant to increase the throughput of global
synchronization.

\section{Implementation}
\label{sec:approach}
Our goal is to use multithreading to tolerate
the latency of random accesses to memory in a global address space
spread across multiple nodes.  Our implementation must provide three features:
\begin{description}
\item[Lightweight multithreading:] We expect to need hundreds of
  threads to tolerate a cluster's network latency, so our threading
  implementation must have low overhead.

\item[Concurrency in global memory references:] A thread making a
  long-latency global memory reference must be able to yield and allow
  other threads to make progress.

\item[Low overhead synchronization:] 
%  We must support global synchronization between threads that allows unrelated threads to proceed unblocked.
  Similarly to a regular memory access, long-latency synchronization
  should also cause a thread to yield and allow others to make
  progress. %\todo{repetitive with the previous item}
\end{description}
We discuss our approach to solving each of these challenges in
turn. For each one, we discuss both our current single-node
implementation and ideas for a multi-node implementation.

@@

%\paragraph{Single-node system.}
In our single-node implementation, we prefetch and yield before performing
a synchronization operation. We support only the synchronization
operations supported by our development platform; we have not
implemented full-bits yet.

@@

Our evaluation consists of two parts. First, we demonstrate that the
runtime system can achieve the same performance as when 
explicit memory concurrency is available. Second, we look at the runtime
within the multi-node picture and show that it can tolerate large
latencies. In each part we begin by studying relevant aspects of our
test machine's memory system so we can interpret our runtime results.

@@

To find the maximum random reference rate that one chip's cores
can issue, we ran pointer chasing code following the model of
Figure~\ref{fig:code}a. Each core issues $n$ list traversals
in a loop; we call $n * \textit{number of cores}$ the number of
{\em concurrent offered references}, since the memory system may not
be able to satisfy them all in parallel. Since our goal is to find a
baseline for evaluating our coroutine library, we depend on the
cores' exploitation of ILP for memory concurrency, rather than our coroutine
library.

@@

The lists were laid out randomly with pointers spaced at a cache line granularity, maximizing
the probability of each reference being a miss. We allocated the lists
on 1GB huge pages to minimize TLB refill overhead.

@@

We varied the size of the per-coroutine local working set from 2KiB
to 4MiB, and ran 0 to 32 local updates for each pointer
traversal. We measured the remote reference rate (i.e. accesses to \texttt{remoteList} from Figure~\ref{fig:code}c).

Figure~\ref{fig:cache-pressure}a shows the result for 8 coroutines per
core.  With 1 to 4 updates per remote reference for smaller working
sets, the runtime achieves near maximum throughput. 
%As working set
%size increases, cache misses in the local space become a
%bottleneck. 
%When there are several updates to large working sets, 
As the number of updates and the working set size increase, the data
of other coroutines is likely to be evicted from cache and performance
suffers.

@@

The results of the experiment with six cores are shown in
Figure~\ref{fig:network-delay}. The runtime can still reach a
maximum rate of \mrps{173}\footnote{The throughput keeps increasing up to around 564 total concurrent references offered (at which point QPI is saturated), far more than the 36 we know the chip to support. This is because we have modeled the network latency by forcing coroutines to wait the extra time before using the data, so the physical miss buffers on our machine are not tied up any longer than usual.}, saturating the QPI link.  Little's law predicts that 300 concurrent references are required to achieve the estimated network rate of \mrps{100} when there is $3\mu s$ latency. The results agree: \mrps{100} is reached with about 53 coroutines per core. Although there will be other overheads in a full runtime, this experiment suggests there is potential to tolerate multi-node system latencies.

Figure~\ref{fig:cache-pressure}b shows the results of the cache pressure experiment for supporting 48 coroutines per core, about the number of contexts needed to tolerate latency in the network experiment. We observe that when performing up to 8 local updates per remote reference, smaller working sets still allow the desired throughput. When performing 1 to 2 updates, the runtime can maintain throughput for working sets of up to 32KiB.

@@

Much work exists on using multithreading to tolerate latency. Hardware
implementations include the Tera MTA~\cite{tera} and Cray
XMT~\cite{feo-xmt}, Simultaneous Multithreading \cite{tullsen-smt},
MIT Alewife \cite{agarwal-alewife}, Cyclops \cite{almasi-cyclops}, and
even GPUs \cite{gpus}. Software-only approaches exist as well; the
Software Controlled Multithreading project \cite{mowry-scm}, QThreads
\cite{qthreads}, MAESTRO \cite{maestro}, and Charm \cite{charm} all
use variants of this idea.
%In contrast to all of these, our goal is to tolerate
%latencies across a multi-node system using mostly software.

@@

We presented our plans for building a system for large graph
computation using commodity parts. We leverage decades-old ideas on
using a large amount of concurrency to tolerate latencies; but we do so
mostly in software. We developed a runtime system for latency
tolerance and our results showed that it is able to manage enough
concurrency to tolerate long latencies to hide the cost of
expensive accesses to our large global memory system (high-latency,
but high-bandwidth).

%\bibliographystyle{acm}
\bibliographystyle{plain}
\renewcommand{\bibfont}{\footnotesize}  
\bibliography{softxmt-hotpar2011}
\end{document}

>>>>

%\textheight 9in
%\textwidth 6.5in
%\columnsep 1pc
%\renewcommand{\baselinestretch}{0}

@@

\authorinfo{Jacob Nelson$^{\dagger}$,
  Brandon Myers$^{\dagger}$,
  A. H. Hunter$^{\dagger}$,
  Preston Briggs$^{\dagger}$
  Luis Ceze$^{\dagger}$,
  Carl Ebeling$^{\dagger}$,\\
  Dan Grossman$^{\dagger}$,
  Simon Kahan$^{{\dagger \ddagger}}$,
  Mark Oskin$^{\dagger}$,
}{\textdagger University of Washington, \textdaggerdbl Pacific Northwest National Laboratory}{
  \{nelson, bdmyers, ahh, preston, luisceze, ebeling, djg, skahan, oskin\}@cs.washington.edu}

@@

\begin{abstract}
   Crunching large graphs is the basis of many emerging applications,
   such as social network analysis and bioinformatics. Graph analytics
   algorithms exhibit little locality and therefore present significant
   performance challenges. Hardware multithreading systems (e.g., Cray
   {\small XMT}) show that with enough concurrency, we can tolerate long
   latencies. Unfortunately, this solution is not available with
   commodity parts.

   Our goal is to develop a latency-tolerant system built out of
   commodity parts and mostly in software. The proposed system includes
   a runtime that supports a large number of lightweight contexts,
   full-bit synchronization and a memory manager that provides a
   high-latency but high-bandwidth global shared memory. This paper
   lays out the vision for our system and justifies its feasibility
   with a performance analysis of the runtime for latency tolerance.

@@

\renewcommand{\baselinestretch}{0.8}

@@

Many important emerging applications such as social network analysis,
bioinformatics, and sensor networks rely on crunching very large
graphs. Unfortunately, the computational cost of these applications
gets quickly out of hand. While tools to analyze social networks and
query semantic graphs with billions of vertices and edges exist
today~\cite{joslyn:hpsf, ediger:msda, brocheler:cosi, pegasus,
  pregel}, graphs of interest to defense applications are expected to
have trillions of entities~\cite{hartley:mssg, kolda:imkd}. Speeding
up these applications at a low cost would have a large impact in how
we analyze large data-sets and make them even more valuable.

@@

Multithreading is a technique that has been used successfully to
implement efficient computations for these
graphs~\cite{bader:bfsmta}. The Cray {\small XMT} is an example of such an
approach: it solves the memory latency problem through concurrency
rather than caching. Each {\small XMT} processor supports 128 hardware contexts
and 1024 outstanding memory operations, and is able to switch contexts
every cycle~\cite{tera, feo-xmt}. This ability comes at a cost: the
{\small XMT} is an expensive, non-commodity machine.

@@

We believe we can build a system based mostly on commodity parts that
can attain {\small XMT}-like performance with a familiar, {\small XMT}-like,
programming model, but at a fraction of the cost. Therefore, our goal
is to build a system that has good performance on low-locality graph
codes but is implemented with cheap commodity processors and simple
additional support in an FPGA. This approach has the added
benefit of not sacrificing general-purpose performance.

@@

Our goal is to preserve the {\small XMT} programming model: large shared global
address space, explicit concurrency with a large number of threads,
and full-bit synchronization.  We partition the overall address
space into a global shared address space and per node private address
spaces. Locality may be exploited directly by the programmer in the
node private address spaces, just as in a conventional cache-coherent
multiprocessor node. In contrast, locality cannot be exploited in the
global space: its value is in providing high bandwidth random access
to any word in a shared data structure by any processor.

@@

The model promises efficiency subject to the presence of sufficient
concurrency.  Exactly how that concurrency is expressed by the
programmer is language dependent. The goal of our system is to support
that concurrency and consequently provide latency tolerance on a large
address space. We do so via software multithreading: computation that
is about to execute a long latency operation (e.g., a remote memory
reference) initiates the operation and then yields to the scheduler,
which quickly resumes execution of another computation. Concurrency is
also used to make more efficient use of network bandwidth: it
enables aggregation of short memory requests into coarser requests,
which leads to large network messages and consequently better
bandwidth utilization.

@@

Synchronization on locations in the private address space of a node
works the the same way as on conventional systems.  Synchronization on
global addresses works differently: we provide atomic operations such
as {\tt int\_fetch\_add} as well as operations using full-bits, as on
the Cray {\small XMT}.  As is true for other long latency operations,
synchronization latency is tolerated by yielding to the scheduler.
Even non-blocking global atomics yield so the core can switch to
another computation while the synchronization is accomplished outside
of the pipeline.  In addition, the system can also exploit aggregation
when concurrency is abundant to increase the throughput of global
synchronization.

\section{Implementation}
\label{sec:approach}
Our goal is to use multithreading to tolerate
the latency of random accesses to memory in a global address space
spread across multiple nodes.  Our implementation must provide three features:
\begin{description}
\item[Lightweight multithreading] We expect to need hundreds of
  threads to tolerate a cluster's network latency, so our threading
  implementation must have low overhead.

\item[Concurrency in global memory references] A thread making a
  long-latency global memory reference must be able to yield and allow
  other threads to make progress.

\item[Low overhead synchronization] 
%  We must support global synchronization between threads that allows unrelated threads to proceed unblocked.
  Similarly to a regular memory access, long-latency synchronization
  should also cause a thread to yield and allow others to make
  progress. %\todo{repetitive with the previous item}
\end{description}
We discuss our approach to solving each of these challenges in
turn. For each one, we discuss both our current single-node
implementation and ideas for a multi-node implementation.

@@

%\paragraph{Single-node system.}
In our single-node implementation, we prefetch and yield before performing
a synchronization operation. We support only the synchronization
operations supported by our development platform; we have not yet
implemented full-bits.

@@

Our evaluation consists of two parts. First, we demonstrate that the
runtime system can achieve the same performance as when 
explicit memory concurrency is available. Second, we look at the runtime
within the multi-node picture and show that it can tolerate large
latencies. In each part, we begin by studying relevant aspects of our
test machine's memory system so we can interpret our runtime results.

@@

To find the maximum random reference rate that one chip's cores
can issue, we ran pointer chasing code following the model of
Figure~\ref{fig:code}a. Each core issues $n$ list traversals
in a loop; we call $n * \textit{number of cores}$ the number of
{\em concurrent offered references} since the memory system may not
be able to satisfy them all in parallel. Since our goal is to find a
baseline for evaluating our coroutine library, we depend on the
cores' exploitation of ILP for memory concurrency, rather than our coroutine
library.

@@

The lists were laid out randomly with pointers spaced at a cache line granularity, maximizing
the probability of each reference being a miss. We allocated the lists
on 1~GB huge pages to minimize TLB refill overhead.

@@

We varied the size of the per-coroutine local working set from 2~KB
to 4~MB, and ran 0 to 32 local updates for each pointer
traversal. We measured the remote reference rate (i.e., accesses to \texttt{remoteList} from Figure~\ref{fig:code}c).

Figure~\ref{fig:cache-pressure}a shows the result for 8~coroutines per
core.  With 1~to 4~updates per remote reference for smaller working
sets, the runtime achieves near maximum throughput. 
%As working set
%size increases, cache misses in the local space become a
%bottleneck. 
%When there are several updates to large working sets, 
As the number of updates and the working set size increase, the data
of other coroutines is likely to be evicted from cache and performance
suffers.

@@

The results of the experiment with six cores are shown in
Figure~\ref{fig:network-delay}. The runtime can still reach a
maximum rate of \mrps{173},\footnote{The throughput keeps increasing
  up to around 564 total concurrent references offered (at which point
  QPI is saturated), far more than the 36 we know the chip to
  support. This is because we have modeled the network latency by
  forcing coroutines to wait the extra time before using the data, so
  the physical miss buffers on our machine are not tied up any longer
  than usual.} saturating the QPI link.  Little's Law predicts that 300 concurrent references are required to achieve the estimated network rate of \mrps{100} when there is $3\mu s$ latency. The results agree: \mrps{100} is reached with about 53 coroutines per core. Although there will be other overheads in a full runtime, this experiment suggests there is potential to tolerate multi-node system latencies.

Figure~\ref{fig:cache-pressure}b shows the results of the cache pressure experiment for supporting 48 coroutines per core, about the number of contexts needed to tolerate latency in the network experiment. We observe that when performing up to 8 local updates per remote reference, smaller working sets still allow the desired throughput. When performing 1 to 2 updates, the runtime can maintain throughput for working sets of up to 32~KB.

@@

Much work exists on using multithreading to tolerate latency. Hardware
implementations include the Tera MTA~\cite{tera} and Cray
{\small XMT}~\cite{feo-xmt}, Simultaneous Multithreading \cite{tullsen-smt},
MIT Alewife \cite{agarwal-alewife}, Cyclops \cite{almasi-cyclops}, and
even GPUs \cite{gpus}. Software-only approaches exist as well; the
Software Controlled Multithreading project \cite{mowry-scm}, QThreads
\cite{qthreads}, MAESTRO \cite{maestro}, and Charm \cite{charm} all
use variants of this idea.
%In contrast to all of these, our goal is to tolerate
%latencies across a multi-node system using mostly software.

@@

We presented our plans for building a system for large graph
computation using commodity parts. We leverage decades-old ideas on
using a large amount of concurrency to tolerate latencies; but we do so
mostly in software. We developed a runtime system for latency
tolerance and our results showed that it is able to manage enough
concurrency to tolerate long latencies to hide the cost of
expensive accesses to our large global memory system (high-latency,
but high-bandwidth).
\newpage
%\bibliographystyle{acm}
\bibliographystyle{plain}
\renewcommand{\bibfont}{\footnotesize}  
\bibliography{softxmt-hotpar2011}
\end{document}
